Approximation as Differentiation

### 2025-07-08 main/approx-as-tangents.tex
[ tex ] fiddling

\begin{enumerate}
\item Even without completeness, in Bounded meet posets and L-posets,
  values and their approximations live in the same set. However, in
  the \emph{total} query language we wish to model in
  \secref{language}, we are not directly interested in the behaviour
  of programs on approximations as we would be for partial programs
  with general recursion. (Moreover, in the light of
  \exref{intervals-and-maxima-elements} it is not clear whether
  approximations for partiality and approximations for stability ought
  to be the same thing. \bob{fwd reference to discussion.}) In
  \exref{intervals-and-maxima-elements}, maximal elements could be
  taken to be the ``proper values''. One idea is to restrict to
  conditionally multiplicative or stable functions that preserve
  maximal elements. However, this idea fails at higher order:
  functions that take maximal elements to maximal elements are not
  themselves maximal elements. We could devise a category of L-posets
  with totality predicates (which would pick out maximal elements at
  first-order) and totality preserving functions, but we prefer a more
  direct method of separating values proper from their approximations
  using the \emph{Category of Families} construction we detail in the
  next section.
\item A more practical reason is that we wish to formalise our
  construction in the proof assistant Agda \cite{agda} in order to get
  an \emph{executable} model of the language in
  \secref{language}. Agda's type theory is both predicative and
  constructive. Predicativity means that it does not have complete
  lattices in the classical sense: for a type $X$ we can only get
  suprema and infinma of families in a lower universe level than
  $X$. This is not necessarily a problem, as \citet{dejong21} shows
  how to develop a large amount of domain theory in a predicative
  setting. However, due to constructivity the lifting construction
  (due to \citet{escardo-knapp}) requires a proof of definedness to
  extract a value. Since we are modelling a total language, we prefer
  to have a model that computes directly.
\end{enumerate}

\bob{For future work: what is the connection between FamCPO(LatGal)
  and Laird's bistable biorders, or Berry bidomains?}

% We could have required that L-posets' principal downsets were complete lattices, which would mean that we would only need to require that forward approximation maps preserve all meets and we could compute the backwards map automatically. However, our goal is a computable theory for \GPS, analogous to automatic differentiation. Requiring completeness to compute backwards maps of approximations would require that complete joins are always computable, which will not be the case for approximations of functions.

% \begin{proposition}
%   L-posets and stable functions form a category. \bob{and the main point is that composition of stability witnesses looks like a chain rule!}
% \end{proposition}

% \bob{L-poset is has products and coproducts, but is not closed. To be closed would require each $\downset{x}$ to be a \emph{complete} lattice. However, this is tricky to implement as we need an infinitary operation to compute the reverse approximation map for function types. See \cite{amadio-curien} for the proof that bounded-complete lattices with stable functions form a cartesian closed category.}

>>>>

\begin{enumerate}
\item Even without completeness, in Bounded meet posets and L-posets,
  values and their approximations live in the same set. However, in
  the \emph{total} query language we wish to model in
  \secref{language}, we are not directly interested in the behaviour
  of programs on approximations as we would be for partial programs
  with general recursion. (Moreover, in the light of
  \exref{intervals-and-maxima-elements} it is not clear whether
  approximations for partiality and approximations for stability ought
  to be the same thing. We discuss this further in
  \secref{conclusion}.) In \exref{intervals-and-maxima-elements},
  maximal elements could be taken to be the ``proper values''. One
  idea is to restrict to conditionally multiplicative or stable
  functions that preserve maximal elements. However, this idea fails
  at higher order: functions that take maximal elements to maximal
  elements are not themselves maximal elements. We could devise a
  category of L-posets with totality predicates (which would pick out
  maximal elements at first-order) and totality preserving functions,
  but we prefer a more direct method of separating values proper from
  their approximations using the \emph{Category of Families}
  construction we detail in the next section.
\item A more practical reason is that we wish to formalise our
  construction in the proof assistant Agda \cite{agda} in order to get
  an \emph{executable} model of the language in
  \secref{language}. Agda's type theory is both predicative and
  constructive. Predicativity means that it does not have complete
  lattices in the classical sense: for a type $X$ we can only get
  suprema and infinma of families in a lower universe level than
  $X$. This is not necessarily a problem, as \citet{dejong21} shows
  how to develop a large amount of domain theory in a predicative
  setting. However, due to constructivity the lifting construction
  (due to \citet{escardo-knapp}) requires a proof of definedness to
  extract a value. Since we are modelling a total language, we prefer
  to have a model that computes directly.
\end{enumerate}

### 2025-07-08 main/conclusion.tex
[ tex ] fiddling

>>>>

\paragraph{More Approximation Domains.} The intervals example is
interesting, but is it useful, and are there more examples like this?
One route might be to follow \cite{edalat-heckmann98} and explore
whether metric spaces (which already provide a native notion of
approximation) can be embedded in $\Fam(\LatGal)$.

\paragraph{Alternative Higher-order Models}

1. $\Fam(\PSh_{\CMon}(\LatGal))$

2. Concrete sheaves over $\Fam(\LatGal)$ analogous to Diffeological spaces?

\paragraph{Categorical Models of Differentation}

\paragraph{Recursion and Partiality.}
1. Despite using stable domain theory as our starting point, we have ignored

2. Conjecture that the category with $(X : Cpo, \partial X : X \to \CLatGal)$, where $\CLatGal$ is the category of complete lattices and galois connections is the right setting. Would need to investigate the connections with dI-domains, bidomains, and bistable bidomains (Laird).

\paragraph{Implementation}

Would be better to have an implementation that worked by translation,
instead of interpreting directly as Agda functions. See work by
Smeding and Vákár on translations. To formalise this, we'd need
to... See also $\Fam(\PSh_{\CMon}(Syn))$?

### 2025-07-08 main/models.tex
[ tex ] fiddling

Unfortunately, neither of the categories $\LatGal$ or $\FinVect$
satisfy the hypotheses of this theorem, because neither of them have
infinite products. We will consider ways to rectify this below
\bob{fwd ref}.

@@

\subsubsection{Discrete Completeness}

@@

\bob{In the conclusion, we have to admit to the possibility that
  $\Fam(CLatGal)$ might be sensible. We could have also used
  PShCMon(LatGal) and there is PShCMon(GraphLang) for
  normalisation. We would have different problems with proving
  correctness however.}

@@

Comparing this conjecture to the previous one, we can see that the
difference between $\CM$, $\Lposet$ and $\Fam(\MeetSLat)$,
$\Fam(\LatGal)$ is that the latter have a stronger separation of
points from tangents, somewhat analogous to the situation with
manifolds. Also note that, contrary to the $\Fam(\FinVect)$
representation of manifolds and differentiable maps, (if this
conjecture holds) we do not throw away information about higher
derivatives. It is retained in the order structure of the tangent
fibres.

>>>>

Unfortunately, neither of the categories $\LatGal$ or $\FinVect$
satisfy the hypotheses of this theorem, because neither of them have
infinite products. We will consider ways to rectify this below in
\secref{fixing-completeness}.

@@

\subsubsection{Discrete Completeness}
\label{sec:fixing-completeness}

@@

% \bob{In the conclusion, we have to admit to the possibility that
%   $\Fam(CLatGal)$ might be sensible. We could have also used
%   PShCMon(LatGal) and there is PShCMon(GraphLang) for
%   normalisation. We would have different problems with proving
%   correctness however.}

@@

Comparing this conjecture to the previous one, we can see that the
difference between $\CM$, $\Lposet$ and $\Fam(\MeetSLat)$,
$\Fam(\LatGal)$ is that the latter have a separation of points from
tangents, somewhat analogous to the situation with manifolds. Also
note that, contrary to the $\Fam(\FinVect)$ representation of
manifolds and differentiable maps, (if this conjecture holds) we do
not throw away information about higher derivatives. It is retained in
the order structure of the tangent fibres.
